Towards planning of infantile cardiac surgery based on simulation

### 2012-02-21 Doc/Miccai/miccai2012.tex
article: shells and method description
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@320 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\newcommand{\myvec}[1]{\ensuremath{\mathbf{#1}}}
\newcommand{\mymat}[1]{\ensuremath{\mathbf{#1}}}
\newcommand{\Vx}{\myvec{x}} % position vector
\newcommand{\Vv}{\myvec{v}} % velocity vector
\newcommand{\Va}{\myvec{a}} % acceleration vector
\newcommand{\Vdv}{\myvec{dv}} % change of velocity vector (unknown in implicit CG, and used in constraint solver)

@@

For these reasons there is a demand for tools helping with the
process of pre-operative planning.

@@

\tomas{What to describe? -- shell elements/FEM/mechanics?}
\tomas{shortly mention: rest shape}

@@

@@

\begin{itemize}

    \item we use shell elements to model blood vessels, dynamic system
        (implicit integration), in Sofa framework \tomas{cite Olivier
        \cite{Comas2010b,Comas2010c} -- shells in corotational environment}
    \item using springs to connect the objects is not good --> generates
        excessive amount of energy
    \item to eliminate this we use one mesh for the simulation and
        different mesh to describe the rest shape and we leave the dynamic
        system to stabilize
    \item to reduce more the generation of excessive energies in the system
        at the beginning of the simulation we repetitively change the rest
        shape (linear interpolation)
\end{itemize}

\section{Conclusion}
\subsection{Verification}

Verification is difficult -- limited or no knowledge of real state after
the surgery (can't put the child under CT/MRI again). \ldots

@@

% Some Vim setup. If it interferes with anyones work feel free to remove
% that.
% vim: et tw=75 fdm=marker spell

>>>>

\newcommand{\myvec}[1]{\ensuremath{\mathbf{#1}}}
\newcommand{\mymat}[1]{\ensuremath{\mathbf{#1}}}
\newcommand{\Vx}{\myvec{x}} % position vector
\newcommand{\Vv}{\myvec{v}} % velocity vector
\newcommand{\Va}{\myvec{a}} % acceleration vector
\newcommand{\Vdv}{\myvec{dv}} % change of velocity vector (unknown in implicit CG, and used in constraint solver)
\newcommand{\deriv}[2]{\frac{\partial #1}{\partial #2}} % Partial derivative

@@

For these reasons there is a demand for tools helping with the
process of pre-operative planning.
\tomas{Usefulness of computer aid --
numerical models:
3D visualisation: Hemminger2005}

@@

\section{Elastic Shell Elements}

To model the blood vessels we employed shell elements in the corotational
environment. The main advantage of shells is their physical accuracy. They
are based on the mechanics of continuum. Also the parameters defining the
parameters of the simulation have clear physical interpretation (unlike for
example for mass-spring model).

The common objection against using shell elements in real-time simulations
is the numerical complexity. While this is technically valid, it is not
good enough reason. Using low resolution mesh we can obtain results with
good visual and physical accuracy sufficient for real-time
simulation.\tomas{isn't this too strong?} Given the bending nature of the
shell one can employ interpolation of the shell surface to improve the
visual mesh presented to the user. \tomas{cite Olivier and our unpublished
results}

A shell element is a combination of two other elements: elastic membrane
and bending plate element. While the membrane element is used to compute
in-plane forces of the shell (like stretching or shearing), the bending
plate elements on the other hand describes the bending properties.

\tomas{add image showing the difference}

\tomas{thin in one dimension}

\subsection{Mechanics of Continuum Preliminaries} % {{{

\tomas{recheck this. btw do we really want this?}

The mathematical formulation follows the steps defined by the mechanics of
continuum.\tomas{ref. to some text book}

For every particle $\myvec{x} \in \Omega$ of the deformable body $\Omega$
the deformation $ \phi : \Omega' \to \Omega $ is a mapping from the rest
shape $\myvec{x}' \in \Omega'$ to it's current shape

$$ \phi(x) = \myvec{x}' + u(x) $$

Where $u(x)$ is the displacement for particle $x$. From the Cauchy's strain
tensor

$$
\epsilon = \frac{1}{2}\left( \nabla^T u + \nabla u \right)
$$

\noindent%
we can compute the stress by applying a constitutive law. In our simulation
we are using linear Hooke's law to keep the numerical formulation simple
enough to be able to compute it in real-time. The stress is computed by

$$
\sigma = \mymat{M}\epsilon
$$

\noindent%
where $\mymat{M}$ is the material matrix. Further from the stress and
strain we can compute the strain energy

$$
E = \int_{\Omega} \epsilon^T \sigma \,\mathrm{d}\Omega
$$

\tomas{now should be discretization}

%}}}

\subsection{shells...}

In this subsection we will briefly summarize how the numerical formulation
of the linear model is derived. For more thorough description we refer the
reader to \tomas{add ref.} Let's define $ u_x, u_y $ to be the two
in-plane displacements and $ u_z $ the off-plane function of deflection. We
define the strain-displacement matrices in the following way. For the
membrane element (using the Cauchy's strain tensor)

\begin{equation}
  \mymat{B_m} = \left[ \begin{matrix}
    \deriv{u_x}{x} \\
    \deriv{u_y}{y} \\
    \deriv{u_x}{y} + \deriv{u_y}{x}
  \end{matrix} \right]
\end{equation}

\noindent%
and for the bending plate element (using the Kirchoff-Love theory)

\begin{equation}
  \mymat{B_b} = \left[ \begin{matrix}
    - z \deriv{^2 u_z}{x^2} \\
    - z \deriv{^2 u_z}{y^2} \\
    - 2z \deriv{^2 u_z}{xy}
  \end{matrix} \right]
\end{equation}

From these we can easily compute the stiffness matrices $K_m$ and $K_b$
(for membrane and bending plate element respectively) by integrating over
the volume of the shell

\begin{equation}
  \mymat{K}_i = \int_V \mymat{B}_i^T \mymat{M} \mymat{B}_i \,\mathrm{d}V
\end{equation}

\noindent%
where $i \in \{\mathrm{b},\mathrm{m}\}$ and $\mymat{M}$ is the material
matrix defined by Hooke's law:

\begin{equation}
  \mymat{M} = \frac{E}{1-\nu^2} \left[ \begin{matrix}
    1   & \nu & 0 \\
    \nu & 1   & 0 \\
    0   & 0   & \frac{1-\nu}{2}
  \end{matrix} \right]
\end{equation}

The only two values defining the matrix value are Poisson's ratio $\nu$ and
the Young's modulus $E$.

\subsection{Co-rotational Formulation}

While the Cauchy's strain used in the formulation of the shell element is
invariant to rigid body translation, it is known to be rotationally
noninvariant. This produces ghost deformations just by rigid rotation. One
way to solve the problem is using Green's strain tensor instead whose
definition is, however, non-linear. Therefore we to overcome the problem we
chose to use the corotational formulation instead.\tomas{add ref;
maybe short description?}

@@

\tomas{clearly state that this is not a complete solution where user points
and clicks -- it's just a proof-of-concept}

In this chapter we will elaborate on the internal aspects of the
simulation.

@@

\begin{itemize}
    \item we use shell elements to model blood vessels, dynamic system
        (implicit integration), in Sofa framework \tomas{cite Olivier
        \cite{Comas2010b,Comas2010c} -- shells in corotational environment}
    \item using springs to connect the objects is not good --> generates
        excessive amount of energy
    \item to eliminate this we use one mesh for the simulation and
        different mesh to describe the rest shape and we leave the dynamic
        system to stabilize
        -- we need to map the nodes between topologies
    \item to reduce more the generation of excessive energies in the system
        at the beginning of the simulation we repetitively change the rest
        shape (linear interpolation)
\end{itemize}

We are aiming for non-interactive simulation without requiring the
surgeon to go through the whole process of the surgery. This introduces several
problems we have to deal with.

The first problem is how to deform the model so that the requested parts
can be sutured together. And second how to simulate the suture.
The simplest solution here would be connecting the respective nodes
together with springs. Such solution, however, has a disadvantage of adding
excessive energies to the system.\tomas{do we need a ref.?}

%\section{Conclusion}
%\subsection{Verification}

Verification is difficult -- limited or no knowledge of real state after
the surgery (too risky to put the child under CT/MRI again).

Therefore we are interested in the geometrical evaluation of the result
i.e. no excessive elongations, expected deformations, independence on
physical parameters (E/nu),\ldots or something like that \ldots

@@

% Some Vim setup. If it interferes with anyones work feel free to remove
% that.
% vim: et sw=2 tw=75 fdm=marker spell
